---
title: "Class 05: Data Visualization with GGPLOT"
author: "Elsa Quillin (PID A69043008)"
format: html
---

Today we are playing with plotting and graphics in R. 

There are lots of ways to make cool figures in R.
There is "base" R graphics (`plot()`, `hist()`, `boxplot()` etc.)

There is also add-on packages, like **ggplot**

```{r}
head(cars)
```
Let's plot this with "base" R

```{r}
plot(cars)
```

```{r}
head(mtcars)
```

Let's plot `mpg` vs `disp`

```{r}
plot(mtcars$mpg, mtcars$disp, pch=16)
```
```{r}
hist(mtcars$mpg)
```

## GGPLOT

The main function in the ggplot2 package is `ggplot()`. First I need to install the *ggplot2** package. I can install any package with the function `install.packages()`

> **N.B.** I never want to run `install.packages()` in my quarto source document!!!

```{r}
library(ggplot2)
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point()
```

```{r}
ggplot(cars) + aes(speed) + geom_histogram()
```

Every ggplot needs at least 3 things: 

- The **data** (given with `ggplot(cars)`)
- The **aesthetic** mapping (given with `aes()`)
- The **geom** (given by `geom_point()`)

> For simple canned plots/graphs "base" R is nearly always faster

### Adding more layers

Let's add a line

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() + geom_line()
```

Let's add a line and a title, subtitle, and caption as well as axis labels

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() + geom_smooth(method="lm", se=FALSE) + 
  labs(title = "Silly Plot", x= "Speed (MPH)", y = "Distance (ft)", caption = "when is tea time anyway???") + 
  theme_bw()
```

## Plot some expression data

Read data file from online URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this wee dataset?

There are `r nrow(genes)` genes in this dataset

There are `r ncol(genes)` columns in this dataset


> Q2. How many "up" regulated genes are there?

```{r}
sum(genes$State == "up")

```

There are `r sum(genes$State == "up")` of genes

```{r}
table(genes$State)
```

```{r}
p <- ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point() +
  scale_color_manual(values=c("blue", "grey", "red"))
```

```{r}
p
```

Silly example of adding labels

```{r}
library(ggrepel)

ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=State, label=Gene) +
  geom_point() +
  scale_color_manual(values=c("blue", "grey", "red")) +
  geom_text_repel(max.overlaps=60) +
  theme_bw()
  
```

## Going Further

Playing with some different layers and the gapminder dataset...

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```

```{r}
tail(gapminder)
```

```{r}
ggplot(gapminder) + 
  aes(gdpPercap, lifeExp, color=continent, size=pop) + 
  geom_point(alpha=0.5) +
  facet_wrap(~continent) +
  theme_bw() +
  scale_size_area(max_size = 10)
```












